---
title: "`match.call` Corner Cases"
author: "Brodie Gaslam"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette

vignette: >
  %\VignetteIndexEntry{match.call Corner Cases}
  %\VignetteEngine{knitr::rmarkdown}
  %\usepackage[utf8]{inputenc}
---

## Overview

This package was developed due to behavior of `match.call` in some very specific
corner cases:

* Attempting to match a call that occurred earlier in the dynamic stack
* For functions that include `...` in their formals
* And are called by functions that forward `...`

This issue cropped up when trying to match calls that happened earlier in the
dynamic call stack to the call to `match.call` (e.g. the grand-parent call).

## An Example of the Problem

```{r}
fun0 <- function(...) {
  fun_gpar <- function(b, ...) {
    fun_par <- function(a)
      match.call(fun_gpar, sys.call(sys.parent()), expand.dots=FALSE)
    fun_par()
  }
  fun_gpar(...)
}
fun0(999, 2 + 2, 3 * pi(), 4)
```

Notice how:

* We lost the first argument (999)
* The second and third arguments both show up as `..X` instead of their expressions
* The second argument is mis-matched to `b` (this isn't obvious from the above,
  but it is mismatched).

Also, note that `sys.call(sys.parent())` is not the same as the default value for
argument `call` of `match.call` even though it looks like it is.  The difference
is in the example above, the expression is evaluated in the parent context,
whereas the default `call` value is evaluated within `match.call`'s context.

## Source of Problem

The reason `match.call` doesn't work properly in the aforementioned circumstances
is that `match.call` always uses the lexical stack of the function that invokes
`match.call` to substitute the `...` argument, irrespective of where the call
you are attempting to match is issued.  In our simple example, `match.call` will
be matching the dots in the **lexical** parent of `fun_par`, instead of in the
**calling** frame of `fun_gpar`.  In this case, `match.call` is matching to the
dots in `fun_gpar <- function(b, ...) {` on the second line instead of the dots
in `fun_gpar(...)` on the 7th line.  These look similar, but are really
completely different.

This mismatch causes the problems described above.  Since `fun_gpar` has a
formal argument `b` in addition to `...`, when we invoke `fun_gpar(...)` the
actual `...` argument inside `fun_gpar` will be the original `...` argument
passed to `fun_gpar`, **less** the explicit formal arguments that are matched.
In this case, `b` positionally matches the 999, so the `...` inside
`fun_gpar` no longer contains 999, but `match.call` uses this 999-less `...` to
match the `fun_gpar(...)` call, not the original `...` in the dynamic parent
frame parent of `fun_gpar`.

This behavior is not unreasonable since `match.call` doesn't provide any way to
give it the correct frame for substituting the dots in.

In addition, when substituting dots, R will replace any non-constant values with
`..X`, etc.  Not entirely clear why this happens, but it is done in this following
snippet of code in `subDots` at lines 1258-1268 in `src/main/unique.c` (R 3.0.2):

```c
for(a = dots, b = rval, i = 1; i <= len; a = CDR(a), b = CDR(b), i++) {
  snprintf(tbuf, 10, "..%d",i);
  SET_TAG(b, TAG(a));
  t = CAR(a);
  while (TYPEOF(t) == PROMSXP)
      t = PREXPR(t);
  if( isSymbol(t) || isLanguage(t) )
      SETCAR(b, mkSYMSXP(mkChar(tbuf), R_UnboundValue));   // <--- HERE
  else
      SETCAR(b, t);
}
```

## More Examples

### Scenario Summary

Here is a comprehensive list of the scenarios that we have found to cause
problems with `match.call`:

1. `fun_gpar` must be a LEXICAL parent of `fun_par` otherwise we get error
   "... used in a situation where it does not exist"; this demonstrates that
   `match.call` descends through the lexical stack, not the dynamic stack.
2. If `fun_par` has a `...` in its formal definition, then none of the
   `...` arguments of `fun_gpar` are captured if the call to `fun_gpar`
   doesn't also include the `...` arguments. This is because we redefine
   `...` to be the arguments to `fun_par`, but then our call to `fun_par`
   doesn't have any arguments, so the lexically earliest set of dots is empty.
3. If `fun_par` doesn't have any formals, then `match.call` will grab
   the `...` values, but with limitations:
     a. unnamed arguments in `...` will be consumed by the named formals of
        `fun_gpar` to the extent those are not otherwise matched, but
       these consumed arguments will not show up in the output of `match.call`
     b. If the call to `fun_gpar` involves expressions or symbols, these are
        replaced with `..1`, `..2`, etc. instead of being captured properly

See below for actual code examples

### Scenario 1:

```{r, error=TRUE, purl=FALSE}
fun1 <- function(a, ...) fun_gpar(a, ...)
fun_gpar <- function(b, ...) fun_par()
fun_par <- function() match.call(fun_gpar, sys.call(sys.parent()), expand.dots=F)

fun1(3, "test", x=45, zest="lemon")
```

### Scenario 2:

```{r}
fun2 <- function(a, ...) {
  fun_gpar <- function(b, ...) {
    fun_par <- function(...) match.call(fun_gpar, sys.call(sys.parent()), expand.dots=F)
    fun_par()
  }
  fun_gpar(a, ...)
}
fun2(3, "test", x=45, zest="lemon")
```

### Scenario 3.a:
```{r}
fun3 <- function(a, ...) {
  fun_gpar <- function(b, c, d, ...) {
    fun_par <- function() match.call(fun_gpar, sys.call(sys.parent()), expand.dots=F)
    fun_par()
  }
  fun_gpar(a, ...)
}
fun3(3, "test", 59, x=45, zest="lemon", 58)
```
We lost "test" and "59", because `...` was reduced by the formals otherwise
not matched (`b` was directly matched by `a`), and after reduction, it was
used to fill the call.

### Scenario 3.b:
```{r}
fun3(3, "test", 59, x=45, zest="lemon", (58), (60))
```

Expressions are returned as their position in dots, since we lost the first two elements we're left with third and fourth.
